---
title: "simul_gorilla_data"
author: "DVM Bishop/P Thompson"
date: "26/02/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(tidyverse)
require(lme4)
require(lmerTest)
options(scipen = 999)
```

## Background

Simulated data from the Treasure Hunt Game to establish power.
Step 1. Simulate a series of Items with a given probability of success that varies by (a) group and (b) item type. Also include estimates of random intercepts for subject and item.

This chunk just uses estimates of effects (in terms of how they affect probability of success) to do a sanity check

```{r makedata}
nItem<-20 #Items per session and condition
nsub <- 25 #subjects per group
ncond <- 2 # 2 conditions (repeated and non)
nsess <- 4 # 4 sessions
ngroup <- 2
simdat <- data.frame(matrix(NA, nrow=nItem*nsub*ncond*nsess*ngroup,ncol=6))
colnames(simdat) <- c('Subject','Group','Session','Item','Condition','Correct')
pcorrsess <- c(.7,.75,.8,.85) #Session effect, estimated as 5% increase in accuracy for each session
p.interaction <- matrix(c(-.075,.075,0,0),nrow=2) #matrix row is condition, column is group.
#These values are added to pcorrsess to define interaction between group and condition
#The sum of the absolute values of first 2 terms defines the difference in conditions for group 1 as a proportion. The sum of absolute values of last 2 terms does the same for group 2 (set to null effect here).
#Thus if group 1 values are +/- .075, then we predict a 15% difference in accuracy between the two conditions for group 1 overall. 
# NB this matrix will also determine the size of any main effect of group and condition

#We are not including any prediction of an interaction with session.

# The following terms for group and condition just refer to the coding of these effects in analysis, and NOT to effect sizes
groupcode <- c(-1,1)
condcode <-c(-1,1)
# These are random effects estimates
itemeff <- (runif(nItem)-.5)/10 #item-specific effect
subeff <- (runif(nsub)-.5)/10 #subject specific effect - these are both small probabilities +ve or -ve
thisrow<-0
for (sess in 1:nsess){
  for (group in 1:ngroup){
     for (sub in 1:nsub){
      for (cond in 1:ncond){
         for(Item in 1:nItem){
       thisrow <- thisrow+1
        simdat$Subject[thisrow]<-sub+1000*group #use large offset for subject code to ensure no overlap in subject IDs for groups 1 and 2. 
        #Group 1 will be numbered 1,2,3, etc and Group 2 1001,1002,1003 etc
         simdat$Group[thisrow]<-groupcode[group]
         simdat$Item[thisrow]<-Item
          simdat$Session[thisrow]<-sess
           simdat$Condition[thisrow]<-condcode[cond]
           criticalp <- pcorrsess[sess]+p.interaction[cond,group]+subeff[sub]+itemeff[Item]
           response<-0
           if(runif(1)<criticalp) 
             {response<-1}
             simdat$Correct[thisrow]<-response
      }
    }
  }
  }
}
#Simdat2 was used in course of checking simulated data: it crunches data to give proportions correct for each Session/Group/subject/condition
#It is not used further in analysis
simdat2 <- aggregate(simdat$Correct, by=list(simdat$Session,simdat$Group,simdat$Subject,simdat$Condition),FUN=mean)
colnames(simdat2)<-c('Session','Group','Subject','Condition','Correct')
```

```{r plotsims}
#Check scores are as intended

myag <- aggregate(simdat$Correct, by=list(simdat$Session,simdat$Group,simdat$Condition),
  FUN=mean)
myagsd <- aggregate(simdat$Correct, by=list(simdat$Session,simdat$Group,simdat$Condition),
  FUN=sd)
myag<-cbind(myag,myagsd[,4]/sqrt(nsub))
colnames(myag)<-c('Session','Group','Condition','p.Corr','se')


group1<-filter(myag,Group==-1,Condition==1)
group2<-filter(myag,Group==1,Condition==1)

plot(group1$Session,group1$p.Corr,type='b',ylim=c(.5,1),xaxt="n",xlab='Session',ylab='percent correct')
axis(side=1, at=1:4, line=1)
lines(group2$Session,group2$p.Corr,type='b',col='red')

group1a<-filter(myag,Group==-1,Condition==-1)
group2a<-filter(myag,Group==1,Condition==-1)
lines(group1a$Session,group1a$p.Corr,type='b',lty=2,col='black')
lines(group2a$Session,group2a$p.Corr,type='b',lty=2,col='red')

# Add a legend
legend('bottomright', legend=c("DLD_repeated","TD_repeated","DLD_novel", "TD_novel"),
       col=c("black", "red","black", "red"), lty=c(1,1,2,2), cex=0.8)


myagg2<-aggregate(myag$p.Corr, by=list(myag$Group,myag$Condition),FUN=mean)
colnames(myagg2)<-c('Group','Condition')
```

# Run statistics on the model


Following advice from here 
https://stats.stackexchange.com/questions/58745/using-lmer-for-repeated-measures-linear-mixed-effect-model

```{r runmodel}

m5 <- lmer(Correct ~ Group*Condition*Session + (1|Subject) + (1|Item), simdat)
summary(m5)

m6 <- lm(Correct~Group*Condition*Session,simdat) #basic Anova without any random effects
summary(m6)

```


```{r power_sim}
data_sims <- function(nItem = 20, nsub = 50, ncond = 2, nsess = 4, ngroup = 2,int_eff=0.075,sess_effect = c(.7,.75,.8,.85))
{
# nItem = Items per session per condition
# nsub = subjects per group
# ncond  = 2 conditions (repeated and non)
# nsess  = 4 sessions
# ngroup = 2 groups
# int_eff = estimate of increase/decrease in p.correct for group 1 in condition 1/2
# sess_effect = estimate of proportion correct for sessions 1,2,3,4 (across all groups/conditions)
  
simdat <- data.frame(matrix(NA, nrow=nItem*nsub*ncond*nsess*ngroup,ncol=6))
colnames(simdat) <- c('Subject','Group','Session','Item','Condition','Correct')
pcorrsess <- sess_effect
p.interaction <- matrix(c(-int_eff,int_eff,0,0),nrow=2) #matrix row is condition, column is group.
#These values are added to pcorrsess to define interaction

#
groupcode <- c(-1,1)
condcode <-c(-1,1)
itemeff <- (runif(nItem)-.5)/10 #item-specific effect
subeff <- (runif(nsub)-.5)/10 #subject specific effect - these are both small probabilities +ve or -ve
thisrow<-0
for (sess in 1:nsess){
  for (group in 1:ngroup){
     for (sub in 1:nsub){
     for (cond in 1:ncond){
         for(Item in 1:nItem){
       thisrow <- thisrow+1
        simdat$Subject[thisrow]<-sub+100*group
         simdat$Group[thisrow]<-groupcode[group]
         simdat$Item[thisrow]<-Item
          simdat$Session[thisrow]<-sess
           simdat$Condition[thisrow]<-condcode[cond]
           criticalp <- pcorrsess[sess]+p.interaction[cond,group]+subeff[sub]+itemeff[Item]
           response<-0
           if(runif(1)<criticalp) 
             {response<-1}
             simdat$Correct[thisrow]<-response
      }
    }
  }
  }
}

return(simdat)
}
```


```{r dopower}
require(rlang)
require(evaluate)

powercalc<-function(N_iters = 100)
{
res<-out<-vector(mode="numeric",length=N_iters)  
set.seed(1)
  for(i in 1:N_iters)
  {
  mydat <- data_sims(nItem = 20, nsub = 25, ncond = 2, nsess = 4, ngroup = 2,int_eff=0.075,sess_effect = c(.7,.75,.8,.85)) 
  res[i] <- summary(lmer(Correct ~ Group*Condition*Session + (1|Subject)+(1|Item), mydat))$coefficients[5,5]
  }

out <- ifelse(res>0.05,0,1)

power = mean(out)  
return(list(power=power,res=res,out=out))  
}

mypower <- powercalc(200)

```

```{r power2}
require(lme4)
require(lmerTest)
require(base)
options(warn=2)

recordWarnings <- function(expr)
{
  warn = NULL
  frame_number <- sys.nframe()
  ans <- withCallingHandlers(expr, warning = function(w)
  {
    assign("warn", w, envir = sys.frame(frame_number))
    invokeRestart("muffleWarning")
  })
  list(ans, warn)
}

powercalc<-function(N_iters = 100)
{
res<-out<-vector(mode="numeric",length=N_iters)  
set.seed(1)
for(i in 1:N_iters)
  {
    if(i %% 10==0) {
      # Print on the screen some message
      cat(paste0("iteration: ", i, "\n"))
   }
  
mydat <- data_sims(nItem = 20, nsub = 25, ncond = 2, nsess = 4, ngroup = 2,int_eff=0.075,sess_effect = c(.7,.75,.8,.85)) 

#error handling case #1
poss.err.mod1 = recordWarnings(lmer(Correct ~ Group*Condition*Session + (1|Subject)+(1|Item), mydat))


res[i] <-ifelse(!is.null(poss.err.mod1[[2]]),NA,summary(lmer(Correct ~ Group*Condition*Session + (1|Subject)+(1|Item), mydat))$coefficients[5,5])

 if(is.na(res[i]))
 {out[i]<-NA}
 else{out[i] <- ifelse(res[i]>0.05,0,1)}

print(paste0('res=',res[i]))
print(paste0('out=',out[i]))
cat("","\n")
}


#    results.file <- tempfile("iteration_", tmpdir=getwd())
#     save(results.0, results.2.5, results.5, results.10, results.20,
#          results.40, results.80, p, file=results.file)
# 
 power = mean(out,na.rm=TRUE)  
 return(list(power=power,res=res,out=out)) 
}

mypower <- powercalc(1000)


```
